---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gdalraster

<!-- badges: start -->
[![R-CMD-check](https://github.com/USDAForestService/gdalraster/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/USDAForestService/gdalraster/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/ctoney/gdalraster/branch/main/graph/badge.svg?token=MXIOPZQ2IU)](https://app.codecov.io/gh/ctoney/gdalraster)
[![CRAN status](https://www.r-pkg.org/badges/version/gdalraster)](https://CRAN.R-project.org/package=gdalraster)
[![cran checks](https://badges.cranchecks.info/worst/gdalraster.svg)](https://cran.r-project.org/web/checks/check_results_gdalraster.html)
[![r-universe status](https://usdaforestservice.r-universe.dev/badges/gdalraster)](https://usdaforestservice.r-universe.dev/gdalraster)
<!-- badges: end -->

## Overview

**gdalraster** is an R interface to the Raster API of the Geospatial Data Abstraction Library ([GDAL](https://gdal.org/)). Calling signatures resemble those of the native C, C++ and Python APIs provided by the GDAL project.

Bindings to GDAL are implemented in the exposed C++ class [`GDALRaster`](https://usdaforestservice.github.io/gdalraster/reference/GDALRaster-class.html) along with several [stand-alone functions](https://usdaforestservice.github.io/gdalraster/reference/index.html#stand-alone-functions), supporting:

* manual creation of uninitialized raster datasets
* creation from existing raster as template
* read/set raster dataset parameters
* low-level I/O
* read/set color tables and raster attribute tables
* copy files/move/rename/delete datasets 
* virtual raster (VRT) for virtual subsetting, resampling and kernel filtering
* `gdalwarp` wrapper for reprojection
* coordinate transformation
* spatial reference convenience functions
* several GDAL algorithms

Additional functionality includes:

* class [`RunningStats`](https://usdaforestservice.github.io/gdalraster/reference/RunningStats-class.html) calculates mean and variance in one pass, and tracks the min, max, sum, and count (i.e., summary statistics on a data stream). The input data values are not stored in memory, so this class can be used to compute statistics for very large data streams.
* class [`CmbTable`](https://usdaforestservice.github.io/gdalraster/reference/CmbTable-class.html) identifies and counts unique combinations of integer values using a hash table.
* [`combine()`](https://usdaforestservice.github.io/gdalraster/reference/combine.html) overlays multiple rasters so that a unique ID is assigned to each unique combination of input values. Pixel counts
for each unique combination are obtained, and combination IDs are optionally written to an output raster.
* [`calc()`](https://usdaforestservice.github.io/gdalraster/reference/calc.html) evaluates an R expression for each pixel in a raster layer or stack of layers. Individual pixel coordinates are available as variables in the R expression, as either x/y in the raster projected coordinate system or inverse projected longitude/latitude.
* [`plot_raster()`](https://usdaforestservice.github.io/gdalraster/reference/plot_raster.html) displays raster data using base R graphics.

**gdalraster** may be suitable for applications that primarily need low-level raster I/O or prefer a direct GDAL API. The additional functionality is somewhat aimed at thematic data analysis but may have other utility.

## Installation

The released version can be installed from CRAN with:

```{r, eval=FALSE}
install.packages("gdalraster")
```

The development version can be installed from [R-universe](https://r-universe.dev/search/). The [R-universe package server](https://ropensci.org/blog/2021/03/04/r-universe-buildsystem/#part-3-a-high-performance-package-server) implements the standard repository format, so packages (incl. Win/Mac binaries) can be installed the same way as from CRAN (no Rtools/devtools/sysreqs required):

```{r, eval=FALSE}
# Install the development version of gdalraster from R-universe
install.packages('gdalraster', repos = c('https://usdaforestservice.r-universe.dev', 'https://cloud.r-project.org'))
```

### From source code

**gdalraster** depends on packages [Rcpp](https://CRAN.R-project.org/package=Rcpp) and [xml2](https://CRAN.R-project.org/package=xml2). These should be installed in R prior to installing **gdalraster** from source.

#### Windows

Installing from source on Windows requires [RTools](https://cran.r-project.org/bin/windows/Rtools/). RTools since version 4.2 includes recent versions of the geospatial libraries (GDAL, GEOS, PROJ) needed to compile **gdalraster**.

With RTools installed, the development version of **gdalraster** can be installed from GitHub with:

```{r, eval=FALSE}
remotes::install_github("USDAForestService/gdalraster")
```

#### Linux

GDAL (>= 2.4.0, built against GEOS), PROJ (>= 4.8.0), and sqlite3 libraries are required. On Ubuntu, recent versions of geospatial libraries can be installed from the [ubuntugis-unstable PPA](https://launchpad.net/~ubuntugis/+archive/ubuntu/ubuntugis-unstable):

```
sudo add-apt-repository ppa:ubuntugis/ubuntugis-unstable
sudo apt update
sudo apt install libgdal-dev libgeos-dev libproj-dev libsqlite3-dev
```

Then in R:

```{r, eval=FALSE}
# Install the released version from CRAN
install.packages("gdalraster")

# Or the development version from GitHub:
remotes::install_github("USDAForestService/gdalraster")
```

#### macOS

GDAL and PROJ can be installed with Homebrew:

```
brew install pkg-config gdal proj
```

Then in R, `configure.args` is needed:

```{r, eval=FALSE}
# Install the released version from CRAN
install.packages("gdalraster", type = "source", configure.args = "--with-proj-lib=$(brew --prefix)/lib/")

# Or the development version from GitHub:
remotes::install_github("USDAForestService/gdalraster", configure.args = "--with-proj-lib=$(brew --prefix)/lib/")
```

## Documentation
* [Reference Manual](https://usdaforestservice.github.io/gdalraster/reference/)
* [Raster API Tutorial](https://usdaforestservice.github.io/gdalraster/articles/raster-api-tutorial.html)
* [Raster Attribute Tables](https://usdaforestservice.github.io/gdalraster/articles/raster-attribute-tables.html)
* [Raster Display](https://usdaforestservice.github.io/gdalraster/articles/raster-display.html)
* [GDAL Block Caching](https://usdaforestservice.github.io/gdalraster/articles/gdal-block-cache.html)
