---
title: "GDAL Config Quick Ref"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{GDAL Config Quick Ref}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Configuration options are essentially global variables the user can set. They are used to alter the default behavior of certain raster format drivers, and in some cases the GDAL core. A large number of configuration options are available. An overall discussion along with a full list of available options and where they apply is in the GDAL documentation at https://gdal.org/user/configoptions.html.

This vignette is a quick reference for a small subset of configuration options that are commonly used, with links to topic-specific documentation provided by the GDAL project. The examples configure from R with `gdalraster::set_config_option()`. The values used in the examples are reasonable but not necessarily recommended as they are context dependent.

```{r}
library(gdalraster)
```

## General options

GDAL doc: https://gdal.org/user/configoptions.html#general-options

**`GDAL_RASTERIO_RESAMPLING`**

The `$read()` method of a `GDALRaster` object will perform automatic resampling if the specified output size (`out_xsize * out_ysize`) is different than the size of the source region being read (`xsize * ysize`). In that case, resampling can be configured to override the default `NEAR` to one of `BILINEAR`, `CUBIC`, `CUBICSPLINE`, `LANCZOS`, `AVERAGE` or `MODE`:

```{r}
set_config_option("GDAL_RASTERIO_RESAMPLING", "BILINEAR")
```

**`CPL_TMPDIR`**

By default, temporary files are written into the current working directory. This can be changed with:

```{r}
set_config_option("CPL_TMPDIR", "<dirname>") # tmpdir to use
```

## Performance and caching

GDAL doc: https://gdal.org/user/configoptions.html#performance-and-caching

**`GDAL_NUM_THREADS`**

Sets the number of worker threads to be used by GDAL operations that support multithreading. This has effect in several different contexts including multi-threaded compression for GeoTiff and multithreaded computation during `warp()` (see respective topics below).

**`GDAL_CACHEMAX`**

The size limit of the block cache is set upon first use (first I/O). Setting `GDAL_CACHEMAX` after that point will not resize the cache. It is a per-session setting. If `GDAL_CACHEMAX` has not been set upon first use of the cache, then the default cache size (`5%` of physical RAM) will be in effect for the current session. See also [GDAL Block Cache](https://usdaforestservice.github.io/gdalraster/articles/gdal-block-cache.html).

```{r}
# set to a specific size in MB
set_config_option("GDAL_CACHEMAX", "800")

# or percent of physical RAM
set_config_option("GDAL_CACHEMAX", "10%")
```

**`GDAL_MAX_DATASET_POOL_SIZE`**

The default number of datasets that can be opened simultaneously by the `GDALProxyPool` mechanism (used by VRT for example) is `100`. This can be increased to get better random I/O performance with VRT mosaics made of numerous underlying raster files. Note: on Linux systems, the number of file handles that can be opened by a process is generally limited to `1024`. This is currently clamped between `2` and `1000`:

```{r}
set_config_option("GDAL_MAX_DATASET_POOL_SIZE", "200") # default is 100
```
Also note that `gdalwarp` [increases the pool size to `450`](https://gdal.org/drivers/raster/vrt.html#performance-considerations) by default.

## Networking

GDAL doc: https://gdal.org/user/configoptions.html#networking-options

## PROJ

GDAL doc: https://gdal.org/user/configoptions.html#proj-options

**`OSR_DEFAULT_AXIS_MAPPING_STRATEGY`**

This option can be set to either `TRADITIONAL_GIS_ORDER` or `AUTHORITY_COMPLIANT`. GDAL >= 3.5 defaults to `AUTHORITY_COMPLIANT`. Determines whether to honor the declared axis mapping of a CRS or override it with the traditional GIS ordering (x = longitude, y = latitude). **Note:** On load, gdalraster sets this option to `TRADITIONAL_GIS_ORDER`.

**`OSR_WKT_FORMAT`**

As of GDAL 3.0, the default format for exporting a spatial reference definition to Well Known Text is WKT 1. This can be overridden with:

```{r}
# SFSQL/WKT1_SIMPLE/WKT1/WKT1_GDAL/WKT1_ESRI/WKT2_2015/WKT2_2018/WKT2/DEFAULT
set_config_option("OSR_WKT_FORMAT", "WKT2")
```

## Warp

GDAL doc: https://gdal.org/programs/gdalwarp.html#memory-usage

The [performance and caching](#performance-and-caching) topic above generally applies to processing with `gdalraster::warp()` (reproject/resample/crop/mosaic).

**`GDAL_NUM_THREADS`**

Multithreaded computation in `warp()` can be enabled with:

```{r}
# note this also affects several other parts of GDAL
set_config_option("GDAL_NUM_THREADS", "4") # number of threads or "ALL_CPUS"
```

Increasing the memory available to `warp()` may also increase performance (i.e., the options passed in `cl_arg` include a value like `c("-wm", "1000")`). The warp memory specified by `"-wm"` is shared among all threads. It is especially beneficial to increase this value when running `warp()` with multithreading enabled.

Multithreading could also be enabled by including a GDAL warp option in `cl_arg` with `c("-wo", "NUM_THREADS=<value>")` greater than 1, which is equivalent to setting the `GDAL_NUM_THREADS` configuration option as shown above.

**`GDAL_CACHEMAX`**

Increasing the size of the I/O block cache may also help. This can be done by setting `GDAL_CACHEMAX` as described in the [performance and caching](#performance-and-caching) topic above.

## GeoTIFF

GDAL doc: https://gdal.org/drivers/raster/gtiff.html#configuration-options

The behavior of the GTiff driver is highly configurable, including with respect to overview creation. For full discussion, see the link above and also the documentation for the [`gdaladdo`](https://gdal.org/programs/gdaladdo.html) command-line utility.

**`GDAL_NUM_THREADS`**

The GTiff driver supports multi-threaded compression (default is compression in the main thread). GDAL documentation states that it is worth it for slow compression algorithms such as `DEFLATE` or `LZMA`. Starting with GDAL 3.6, this option also enables multi-threaded decoding when read requests intersect several tiles/strips:

```{r}
# specify the number of worker threads or ALL_CPUS
# note this also affects several other parts of GDAL
set_config_option("GDAL_NUM_THREADS", "ALL_CPUS")
```

**`COMPRESS_OVERVIEW`**

Raster overviews (also known as pyramids) can be built with the `$buildOverviews()` method of a `GDALRaster` object. It is generally desirable to compress the overviews when building:

```{r}
# applies to external overviews (.ovr), and internal overviews if GDAL >= 3.6
# LZW is a good default but several other compression algorithms are available
set_config_option("COMPRESS_OVERVIEW", "LZW")
```

## AWS S3 buckets

GDAL doc: [/vsis3/](https://gdal.org/user/virtual_file_systems.html#vsis3-aws-s3-files) (AWS S3 file system handler)

**`AWS_NO_SIGN_REQUEST`**

Request signing can be disabled for public buckets that do not require an AWS account (`--no-sign-request` in AWS CLI):

```{r}
# public bucket no AWS account required
set_config_option("AWS_NO_SIGN_REQUEST", "YES")
```

**`AWS_ACCESS_KEY_ID`**  
**`AWS_SECRET_ACCESS_KEY`**  
**`AWS_SESSION_TOKEN`**  
**`AWS_REQUEST_PAYER`**

If authentication is required, configure credentials with:

```{r}
set_config_option("AWS_ACCESS_KEY_ID", "<value>") # key ID
set_config_option("AWS_SECRET_ACCESS_KEY", "<value>") # secret access key
# used for validation if using temporary credentials:
set_config_option("AWS_SESSION_TOKEN", "<value>") # session token
# if requester pays:
set_config_option("AWS_REQUEST_PAYER", "<value>") # requester
```

## Google Cloud Storage

GDAL doc: [/vsigs/](https://gdal.org/user/virtual_file_systems.html#vsigs-google-cloud-storage-files) (Google Cloud Storage files)

## Microsoft Azure

GDAL doc: [/vsiaz/](https://gdal.org/user/virtual_file_systems.html#vsiaz-microsoft-azure-blob-files) (Microsoft Azure Blob files)
